---
title: "team.Rmd"
author: 'Jessie Bustin Github: jebustin'
date: "3/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
defense <- read_excel('cyclonesFootball2019.xlsx', sheet='Defensive')
offense <- read_excel('cyclonesFootball2019.xlsx', sheet='Offensive')
bio <- read_excel('cyclonesFootball2019.xlsx', sheet='Biography')
```

### Part One
### 1.
```{r}
offense$Name = as.factor(offense$Name)
offense$Opponent_Opponent = as.factor(offense$Opponent_Opponent)
defense$Name = as.factor(defense$Name)
defense$Opponent_Opponent = as.factor(defense$Opponent_Opponent)
bio$Name = as.factor(bio$Name)
```

### 2.
```{r}
defClean = defense %>%
  mutate_if(is.character, as.numeric)
offClean = offense %>%
  mutate_if(is.character, as.numeric)
bioClean = bio %>%
  mutate_at("Weight", as.numeric)
```

### 3.
```{r}
bio <- bio %>%
  separate('Height', c("Feet", "Inches"), 
           sep = "-(?=[^ ]+$)", remove = FALSE) %>%
  mutate_at("Feet", as.numeric) %>%
  mutate_at("Inches", as.numeric)

bio$Height <- bio$Feet * 12 + bio$Inches

bio <- bio %>% select(-Feet, -Inches)
```

```{r}
str(defClean)
str(offClean)
str(bioClean)
```

### Part Two(1)
### 1.
```{r}
defCleanTidy <- defClean %>% 
  pivot_longer(Tackles_Solo:Pass_PB, names_to = 'Statistic',
               values_to='Value')
head(defCleanTidy)
```

### 2.
```{r}
defCleanTidy %>% 
  ggplot(aes(x = defCleanTidy$Value)) +
  geom_histogram() + 
  facet_wrap(~Statistic) + 
  ggtitle("Distributions of Defensive Statistics") +
  xlab("Number of Instances")
```

According to the above graphs turnovers of all types are very rare and so are sacks.  This fits with general football knowledge.

### 3.
```{r}
WVandKS <- defCleanTidy %>% 
  filter(Statistic == 'Tackles_Solo') %>% 
  filter(Opponent_Opponent == 'West Virginia' | Opponent_Opponent ==
         'Kansas State') %>%
  pivot_wider(names_from='Opponent_Opponent', 
              values_from='Value')

ggplot(data = WVandKS, aes(x = WVandKS$'West Virginia', y = WVandKS$'Kansas State')) +
  geom_point() +
  ggtitle("Solo Tackles")
```

Iowa State had better defense against Kansas State than they did agaisnt West Virginia in 2019 based on the number of solo tackles.  The above graph helps to show this because there are more points higher on the y axis than there are higher on the x axis.  Let's verify this numerically:

```{r}
defCleanTidy %>% 
  filter(Statistic == 'Tackles_Solo') %>% 
  filter(Opponent_Opponent == 'West Virginia' | Opponent_Opponent ==
         'Kansas State') %>%
  group_by(Opponent_Opponent) %>%
  summarize(Total = sum(Value))
```

Our numerical analysis agrees.


### Part Two(2)
### 1.
```{r}
bioCleanTidy <- bioClean %>%
  separate('Hometown', c("Home_City", "Home_State"), 
           sep = ", (?=[^ ]+$)")

bioCleanTidy %>% head(3)
```

### 2.
```{r}
bioCleanTidy %>%
  group_by(Home_State) %>%
  summarize(Count = n())
```

The above table shows the number of players from each state.

### Part Three
### 1.
```{r}
offBioMerge <- right_join(offClean, bioCleanTidy, by = "Name")
```

### 2.
```{r}
offBioMerge %>% 
  ggplot(aes(x = Weight, y = Receiving_YDS)) +
  geom_point() +
  ggtitle("Player Weight and Receving Yards") +
  xlab("Weight") +
  ylab("Game Yards")
```

There does appear to be a slight correlation between a players weight and their receiving yards.  This relationship is moderate, negative, and not linear.  There does seem to be a player that breaks with this trend as he is on the heavier end and still has higher receiving yards.  This analysis also fits with general knowledge, lighter players are faster and complete more receiving yards with some exceptions like some tight ends.

### 3.
```{r}

```

### 4.
Do in state or out of state players perform better offensively?  Is the same true for the defense?

To answer this question we will add a column to the bio data that is a boolean set to true if the player is from Iowa.  We will then merge offense with bio and defense with bio.  We will then compare offensive receiving yards for Iowa players vs out of state for the season and then total solo tackles for in and out of state players for the season.
```{r}
bioCleanTidy$Iowan <- bioCleanTidy$Home_State == 'Iowa'

offBioMerge <- right_join(offClean, bioCleanTidy, by = "Name")

defBioMerge <- right_join(defClean, bioCleanTidy, by = "Name")

offBioMerge <- offBioMerge %>% 
  mutate(Rushing_YDS = replace_na(Rushing_YDS, 0)) %>%
  mutate(Passing_YDS = replace_na(Passing_YDS, 0)) %>%
  mutate(Receiving_YDS = replace_na(Receiving_YDS, 0)) %>%
  select(Name, Iowan, Rushing_YDS, Passing_YDS, Receiving_YDS) %>%
  mutate(sum = rowSums(.[3:5]))

offBioMerge %>%
  group_by(Iowan, Name) %>%
  summarize(Total_YDS = sum(sum)) %>%
  subset(Total_YDS != 0) %>%
  ggplot(aes(x = Iowan, y = Total_YDS)) +
  geom_boxplot() +
  ggtitle("Offensive Yards By In or Out of State")

offBioMerge %>%
  subset(sum != 0) %>%
  group_by(Iowan) %>%
  summarize(Mean_Player_YDS = sum(sum) / n()) %>%
  print()

offBioMerge %>%
  subset(sum != 0 & Name != 'Purdy, Brock') %>%
  group_by(Iowan) %>%
  summarize(Mean_Player_YDS = sum(sum) / n(), count = n()) %>%
  print()

```

It was decided that if a player had 0 total offensive yards they would not be included as they were not playing or were not in a position that accumulates yardage.  In addition the the graphic representation we figured the average offensive yards per game for in state verses out of state players.  The out of state players perform much better.  To check to see how much the QB, Brock Purdy influenced this number we also removed him and found that out of state players still perform much better offensively.  

```{r}
defBioMerge %>%
  group_by(Iowan, Name) %>%
  subset(Tackles_Solo != 0) %>%
  summarize(Total_Solo_Tackles = sum(Tackles_Solo)) %>%
  ggplot(aes(x = Iowan, y = Total_Solo_Tackles)) +
  geom_boxplot() +
  ggtitle("Solo Tackles by in or Out of State")

defBioMerge %>%
  group_by(Iowan) %>%
  subset(Tackles_Solo != 0) %>%
  summarize(Mean_Player_Tackles = sum(Tackles_Solo, na.rm = TRUE) /
              n(), count = n()) %>%
  print()
```

It appears that out of state players are better on the defensive side of the ball as well but the difference is much closer than on offense.